Local Type Inference for Polarised System F with Existentials

### 2023-09-28 oopsla24/tex/03-declarative-system.tex
minor

\UseRawInputEncoding
% vim: ft=tex
\documentclass[a4,natbib=false]{article}
\usepackage[a4paper, total={8in, 10in}]{geometry}
\usepackage{hyperref}
\usepackage{mathpartir}

\input{prelude.tex}

\newcommand{\ottDir}{../ott/_gen}
\newcommand{\genDir}{_gen}
\input{\ottDir/included.tex}

% \renewcommand{\ottdruleEOneNVarName}[0]{ Hello \def\@currentlabelname{FOO} \phantomsection  }
% \newcommand{\ottdrulename}[1]{\textsc{#1}}

% \renewcommand{\ottdrule}[4][]{{\displaystyle\frac{\begin{array}{l}#2\end{array}}{#3}\quad\ottdrulename{#4}%
%   }\ruleLabel{#4}{#4}}

% \renewcommand{\ottdrule}[4][]{%
%   {\displaystyle\frac{\begin{array}{l}#2\end{array}}{#3}\quad\ottdrulename{#4}}%
%   \mpr@label{\textsc{#1}}%
% }
% \DeclareDocumentCommand \redrule {m m o}{%
%   \inferrule*[vcenter,left={#3:}]{}{#1 \tred #2}
%   \mpr@label{\textsc{#3}}
% }

% \renewcommand{\ottdrulename}[2][#1]{\textsc{#1} \label{}}

% \renewcommand{\ottdruleEOneNVarName}[0]{foo }

% \makeatletter
% \renewcommand{\ottdruleEOneNVar}[1]{\ottdrule[#1]{%
%   }{
%     % \protected@edef\@currentlabelname{foo}
%     \def\@currentlabelname{foo}%
%     \phantomsection
%     \label{boo}
%     \alpha ^{-}   \eqEOne   \alpha ^{-}
%   }{%
%     {\ottdruleEOneNVarName}{}%
%   } }
% \makeatother

\newcommand{\depth}[1]{\ensuremath{\mathsf{depth}(#1)}}
\newcommand{\size}[1]{\ensuremath{\mathsf{size}(#1)}}

\newcommand{\ruleref}[1]{Rule \nameref{#1}}

% ord varset in uN = varset'

\renewcommand{\ottdruleONVarInName}[0]{(Var$_{-\in}^{\text{Ord}}$)}
\renewcommand{\ottdruleONVarNInName}[0]{(Var$_{-\notin}^{\text{Ord}}$)}
\renewcommand{\ottdruleONUVarName}[0]{(UVar$_{-} ^{\text{Ord}}$)}
\renewcommand{\ottdruleOShiftUName}[0]{($\uparrow^{\text{Ord}}$)}
\renewcommand{\ottdruleOArrowName}[0]{($\rightarrow^{\text{Ord}}$)}
\renewcommand{\ottdruleOForallName}[0]{($\forall^{\text{Ord}}$)}

% ord varset in uP = varset'

\renewcommand{\ottdruleOPVarInName}[0]{(Var$_{+\in}^{\text{Ord}}$)}
\renewcommand{\ottdruleOPVarNInName}[0]{(Var$_{+\notin}^{\text{Ord}}$)}
\renewcommand{\ottdruleOPUVarName}[0]{(UVar$_{+}^{\text{Ord}}$)}
\renewcommand{\ottdruleOShiftDName}[0]{($\downarrow^{\text{Ord}}$)}
\renewcommand{\ottdruleOExistsName}[0]{($\exists^{\text{Ord}}$)}

% nf(N) = M
\renewcommand{\ottdruleNrmNVarName}[0]{(Var$_{-}^{\text{nf}}$)}
\renewcommand{\ottdruleNrmNUVarName}[0]{(UVar$_{-}^{\text{nf}}$)}
\renewcommand{\ottdruleNrmShiftUName}[0]{($\uparrow^{\text{nf}}$)}
\renewcommand{\ottdruleNrmArrowName}[0]{($\rightarrow^{\text{nf}}$)}
\renewcommand{\ottdruleNrmForallName}[0]{($\forall^{\text{nf}}$)}

% nf(P) = Q
\renewcommand{\ottdruleNrmPVarName}[0]{(Var$_{-}^{\text{nf}}$)}
\renewcommand{\ottdruleNrmPUVarName}[0]{(UVar$_{-}^{\text{nf}}$)}
\renewcommand{\ottdruleNrmShiftDName}[0]{($\downarrow^{\text{nf}}$)}
\renewcommand{\ottdruleNrmExistsName}[0]{($\exists^{\text{nf}}$)}

% N ≈ M

\renewcommand{\ottdruleEOneNVarName}[0]{(Var$_-^{\eqEOne}$)}
\renewcommand{\ottdruleEOneShiftUName}[0]{($\uparrow^{\eqEOne}$)}
\renewcommand{\ottdruleEOneArrowName}[0]{($\rightarrow^{\eqEOne}$)}
\renewcommand{\ottdruleEOneForallName}[0]{($\forall^{\eqEOne}$)}

% P ≈ Q
\renewcommand{\ottdruleEOnePVarName}[0]{(Var$_+^{\eqEOne}$)}
\renewcommand{\ottdruleEOneShiftDName}[0]{($\downarrow^{\eqEOne}$)}
\renewcommand{\ottdruleEOneExistsName}[0]{($\exists^{\eqEOne}$)}

% G ⊢ N ≤1 M

\renewcommand{\ottdruleDOneNVarName}[0]{(Var$_-^{\subDOne}$)}
\renewcommand{\ottdruleDOneShiftUName}[0]{($\uparrow^{\subDOne}$)}
\renewcommand{\ottdruleDOneArrowName}[0]{($\rightarrow^{\subDOne}$)}
\renewcommand{\ottdruleDOneForallName}[0]{($\forall^{\subDOne}$)}

% G ⊢ P ≥1 Q
\renewcommand{\ottdruleDOnePVarName}[0]{(Var$_+^{\supDOne}$)}
\renewcommand{\ottdruleDOneShiftDName}[0]{($\downarrow^{\supDOne}$)}
\renewcommand{\ottdruleDOneExistsName}[0]{($\exists^{\supDOne}$)}

% G ⊢ N ≈1 M
\renewcommand{\ottdruleDOneNDefName}[0]{($\eqDOne_{-}$)}

% G ⊢ P ≈1 Q
\renewcommand{\ottdruleDOnePDefName}[0]{($\eqDOne_{+}$)}

% G ⊨ iP1 ∨ iP2 = iQ
\renewcommand{\ottdruleLUBVarName}[0]{(Var$^{\vee}$)}
\renewcommand{\ottdruleLUBShiftName}[0]{($\downarrow^{\vee}$)}
\renewcommand{\ottdruleLUBExistsName}[0]{($\exists^{\vee}$)}
\renewcommand{\ottdruleLUBUpgradeName}[0]{(Upg)}

% G ; Θ ⊨ uN ≤ iM ⫤ us
\renewcommand{\ottdruleANVarName}[0]{(Var$_-^{\subA}$)}
\renewcommand{\ottdruleAShiftUName}[0]{($\uparrow^{\subA}$)}
\renewcommand{\ottdruleAArrowName}[0]{($\rightarrow^{\subA}$)}
\renewcommand{\ottdruleAForallName}[0]{($\forall^{\subA}$)}

% G ; Θ ⊨ iP ≥ uQ ⫤ us
\renewcommand{\ottdruleAPVarName}[0]{(Var$_+^{\supA}$)}
\renewcommand{\ottdruleAShiftDName}[0]{($\downarrow^{\supA}$)}
\renewcommand{\ottdruleAExistsName}[0]{($\exists^{\supA}$)}
\renewcommand{\ottdruleAPUVarName}[0]{(UVar$^{\supA}$)}

% Γ ⊢ scE1 & scE2 = scE3 :: :: E :: 'E'
\renewcommand{\ottdruleSCMESupSupName}[0]{$([[≥]]\&^{+}[[≥]])$}
\renewcommand{\ottdruleSCMEEqSupName}[0]{$([[≈]]\&^{+}[[≥]])$}
\renewcommand{\ottdruleSCMESupEqName}[0]{$([[≥]]\&^{+}[[≈]])$}
\renewcommand{\ottdruleSCMEPEqEqName}[0]{$([[≈]]\&^{+}[[≈]])$}
\renewcommand{\ottdruleSCMENEqEqName}[0]{$([[≈]]\&^{-}[[≈]])$}

% Γ ; Θ ⊨ uN ≈u iM ⫤ UC 
\renewcommand{\ottdruleUNVarName}[0]{(Var$_{-}^{[[≈u]]}$)}
\renewcommand{\ottdruleUShiftUName}[0]{($\uparrow^{[[≈u]]}$)}
\renewcommand{\ottdruleUArrowName}[0]{($\rightarrow^{[[≈u]]}$)}
\renewcommand{\ottdruleUForallName}[0]{($\forall^{[[≈u]]}$)}
\renewcommand{\ottdruleUNUVarName}[0]{(UVar$_{-}^{[[≈u]]}$)}

% Γ ; Θ ⊨ uP ≈u iQ ⫤ UC 
\renewcommand{\ottdruleUPVarName}[0]{(Var$_{+}^{[[≈u]]}$)}
\renewcommand{\ottdruleUShiftDName}[0]{($\downarrow^{[[≈u]]}$)}
\renewcommand{\ottdruleUExistsName}[0]{($\exists^{[[≈u]]}$)}
\renewcommand{\ottdruleUPUVarName}[0]{(UVar$_{+}^{[[≈u]]}$)}

% G ⊨ iP1 ≈au iP2 ⫤ ( Ξ , uQ , aus1 , aus2 )
\renewcommand{\ottdruleAUPVarName}[0]{(Var$_{+}^{[[≈au]]}$)}
\renewcommand{\ottdruleAUShiftDName}[0]{($\downarrow^{[[≈au]]}$)}
\renewcommand{\ottdruleAUExistsName}[0]{($\exists^{[[≈au]]}$)}

% G ⊨ iN1 ≈au iN2 ⫤ ( Ξ , uM , aus1 , aus2 )
\renewcommand{\ottdruleAUNVarName}[0]{(Var$_{-}^{[[≈au]]}$)}
\renewcommand{\ottdruleAUShiftUName}[0]{($\uparrow^{[[≈au]]}$)}
\renewcommand{\ottdruleAUForallName}[0]{($\forall^{[[≈au]]}$)}
\renewcommand{\ottdruleAUArrowName}[0]{($\rightarrow^{[[≈au]]}$)}
\renewcommand{\ottdruleAUAUName}[0]{(AU)}

% Γ ⊢ iN 
\renewcommand{\ottdruleWFTNVarName}[0]{(Var$_{-}^{\text{WF}}$)}
\renewcommand{\ottdruleWFTShiftUName}[0]{($\uparrow^{\text{WF}}$)}
\renewcommand{\ottdruleWFTArrowName}[0]{($\rightarrow^{\text{WF}}$)}
\renewcommand{\ottdruleWFTForallName}[0]{($\forall^{\text{WF}}$)}

% Γ ⊢ iP
\renewcommand{\ottdruleWFTPVarName}[0]{(Var$_{+}^{\text{WF}}$)}
\renewcommand{\ottdruleWFTShiftDName}[0]{($\downarrow^{\text{WF}}$)}
\renewcommand{\ottdruleWFTExistsName}[0]{($\exists^{\text{WF}}$)}

% Γ ; Ξ ⊢ uN
\renewcommand{\ottdruleWFATNVarName}[0]{(Var$_{-}^{\text{WF}}$)}
\renewcommand{\ottdruleWFATNUVarName}[0]{(UVar$_{-}^{\text{WF}}$)}
\renewcommand{\ottdruleWFATShiftUName}[0]{($\uparrow^{\text{WF}}$)}
\renewcommand{\ottdruleWFATArrowName}[0]{($\rightarrow^{\text{WF}}$)}
\renewcommand{\ottdruleWFATForallName}[0]{($\forall^{\text{WF}}$)}

% Γ ; Ξ ⊢ uP
\renewcommand{\ottdruleWFATPVarName}[0]{(Var$_{+}^{\text{WF}}$)}
\renewcommand{\ottdruleWFATPUVarName}[0]{(UVar$_{+}^{\text{WF}}$)}
\renewcommand{\ottdruleWFATShiftDName}[0]{($\downarrow^{\text{WF}}$)}
\renewcommand{\ottdruleWFATExistsName}[0]{($\exists^{\text{WF}}$)}

% Γ ⊢ iP : scE
\renewcommand{\ottdruleSATSCESupName}[0]{($[[:≥]]_{+}^\text{sat}$)}
\renewcommand{\ottdruleSATSCEPEqName}[0]{($[[:≈]]_{+}^\text{sat}$)}

% Γ ⊢ iN : scE
\renewcommand{\ottdruleSATSCENEqName}[0]{($[[:≈]]_{-}^\text{sat}$)}

% Γ ; Φ ⊢ v : iP 
\renewcommand{\ottdruleDTVarName}[0]{(Var$^{\text{inf}}$)}
\renewcommand{\ottdruleDTThunkName}[0]{($\{\}^{\text{inf}}$)}
\renewcommand{\ottdruleDTPAnnotName}[0]{(ann$_+^{\text{inf}}$)}
\renewcommand{\ottdruleDTPEquivName}[0]{($[[≈]]_+^{\text{inf}}$)}

% Γ ; Φ ⊢ c : iN 
\renewcommand{\ottdruleDTtLamName}[0]{($\lambda^{\text{inf}}$)}
\renewcommand{\ottdruleDTTLamName}[0]{($\Lambda^{\text{inf}}$)}
\renewcommand{\ottdruleDTReturnName}[0]{(ret$^{\text{inf}}$)}
\renewcommand{\ottdruleDTVarLetName}[0]{(let$^{\text{inf}}$)}
\renewcommand{\ottdruleDTAppLetName}[0]{(let$_@^{\text{inf}}$)}
\renewcommand{\ottdruleDTAppLetAnnName}[0]{(let$_{:@}^{\text{inf}}$)}
\renewcommand{\ottdruleDTUnpackName}[0]{(let$_{\exists}^{\text{inf}}$)}
\renewcommand{\ottdruleDTNAnnotName}[0]{(ann$_-^{\text{inf}}$)}
\renewcommand{\ottdruleDTNEquivName}[0]{($[[≈]]_-^{\text{inf}}$)}

% Γ ; Φ ⊢ iN ● args ⇒> iM 
\renewcommand{\ottdruleDTEmptyAppName}[0]{($\emptyset_{[[●]][[⇒>]]}^{\text{inf}}$)}
\renewcommand{\ottdruleDTArrowAppName}[0]{($\rightarrow_{[[●]][[⇒>]]}^{\text{inf}}$)}
\renewcommand{\ottdruleDTForallAppName}[0]{($\forall_{[[●]][[⇒>]]}^{\text{inf}}$)}

% Γ ; Φ ⊨ v : iP  
\renewcommand{\ottdruleATVarName}[0]{(Var$^{\text{inf}}$)}
\renewcommand{\ottdruleATThunkName}[0]{($\{\}^{\text{inf}}$)}
\renewcommand{\ottdruleATPAnnotName}[0]{(ann$_+^{\text{inf}}$)}

% Γ ; Φ ⊨ c : iN 
\renewcommand{\ottdruleATtLamName}[0]{($\lambda^{\text{inf}}$)}
\renewcommand{\ottdruleATTLamName}[0]{($\Lambda^{\text{inf}}$)}
\renewcommand{\ottdruleATReturnName}[0]{(ret$^{\text{inf}}$)}
\renewcommand{\ottdruleATVarLetName}[0]{(let$^{\text{inf}}$)}
\renewcommand{\ottdruleATAppLetName}[0]{(let$_@^{\text{inf}}$)}
\renewcommand{\ottdruleATAppLetAnnName}[0]{(let$_{:@}^{\text{inf}}$)}
\renewcommand{\ottdruleATUnpackName}[0]{(let$_{\exists}^{\text{inf}}$)}
\renewcommand{\ottdruleATNAnnotName}[0]{(ann$_-^{\text{inf}}$)}

% Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC 
\renewcommand{\ottdruleATEmptyAppName}[0]{($\emptyset_{[[●]][[⇒>]]}^{\text{inf}}$)}
\renewcommand{\ottdruleATArrowAppName}[0]{($\rightarrow_{[[●]][[⇒>]]}^{\text{inf}}$)}
\renewcommand{\ottdruleATForallAppName}[0]{($\forall_{[[●]][[⇒>]]}^{\text{inf}}$)}

% scE1 singular with iP
\renewcommand{\ottdruleSINGPEqName}[0]{($[[≈]]_{+}^{\text{sing}}$)}
\renewcommand{\ottdruleSINGSupVarName}[0]{($[[:≥]]\alpha^{\text{sing}}$)}
\renewcommand{\ottdruleSINGSupShiftName}[0]{($[[:≥]][[↓]]^{\text{sing}}$)}

% scE1 singular with iN
\renewcommand{\ottdruleSINGNEqName}[0]{($[[≈]]_{-}^{\text{sing}}$)}

\begin{document}

\tableofcontents

\newpage

\section{Declarative Type Systems}

\subsection{Grammar}
We assume that there is an infinite set of positive and 
negative \emph{type} variables. Positive type variables are denoted as 
$[[α⁺]]$, $[[β⁺]]$, $[[γ⁺]]$, etc.
Negative type variables are denoted as $[[α⁻]]$, $[[β⁻]]$, $[[γ⁻]]$, etc.
We assume there is an infinite set of \emph{term} variables,
which are denoted as $[[x]]$, $[[y]]$, $[[z]]$, etc.
A list of objects (variables, types or terms) is denoted by
an overline arrow. For instance, $[[pas]]$ is a list of positive type variables, 
$[[nbs]]$ is a list of negative type variables, 
$[[args]]$ is a list of values, which are arguments of a function.
$[[fv(iP)]]$ and $[[fv(iN)]]$ denote the set of free variables 
in a type $[[iP]]$ and $[[iN]]$, respectively.

\bigskip

% \ottgrammartabular{
%   \ottiP\ottinterrule
%   \ottiN\ottinterrule
% }

@@

\subsection{Equalities}
For simplicity, we assume alpha-equivalent terms equal.
This way, we assume that substitutions do not capture bound variables.
Besides, we equate
$[[∀pas.∀pbs.iN]]$ with $[[∀pas,pbs.iN]]$, 
as well as $[[∃nas.∃nbs.iP]]$ with $[[∃nas,nbs.iP]]$,
and lift these equations transitively and congruently 
to the whole system.

\subsection{Contexts and Well-formedness}

\begin{definition}[Declarative Type Context]
  \hfill \\
  Declarative type context $[[Γ]]$ is represented by a set of 
  type variables. The concatenation $[[Γ1, Γ2]]$ means the 
  union of two contexts $[[Γ1 ∪ Γ2]]$.
\end{definition}

$[[Γ ⊢ iP]]$ and $[[Γ ⊢ iN]]$ denote that the type is well-formed in the context $[[Γ]]$,
which, in fact, means that each free type variable of the type is contained in $[[Γ]]$
(it will be shown later in \cref{lemma:wf-soundness,lemma:wf-ctxt-equiv}).

Notice that checking the well-formedness of a type
is an \emph{algorithmic} procedure, in which 
both the context and the type are considered inputs.
In other words, it is syntax-directed and mode-correct 
(according to \cite{dunfieldBidirectionalTyping2020}), 
which means that

\begin{algorithm}[Type Well-formedness]
  \label{alg:wf}
  \hfill

  \begin{multicols}{2}
  \ottdefnWFTNiWF{}
  \columnbreak

  \ottdefnWFTPiWF{}
  \end{multicols}

\end{algorithm}

\subsection{Substitutions}

\begin{definition}[Substitution]
  Substitutions (denoted as $[[σ]]$) 
  are represented by total functions form variables to types, preserving the polarity. 
\end{definition}

\begin{algorithm}[Substitution Application]
  Substitution application is denoted as $[[ [σ]iP ]]$ and $[[ [σ]iN ]]$.
  It is defined naturally as follows:
    \begin{multicols}{2}
      \begin{itemize}
        \item[] $[[ [σ]α⁺ ]] = [[σ]] ([[α⁺]])$;
        \item[] $[[ [σ]α⁻ ]] = [[σ]] ([[α⁻]])$;
        \item[] $[[ [σ]↓iN ]] = [[↓[σ]iN]]$;
        \item[] $[[ [σ]↑iP ]] = [[↑[σ]iP]]$;
        \item[] $[[ [σ](iP → iN) ]] = [[ [σ]iP → [σ]iN ]]$.
        \item[] $[[ [σ]∃nas.iQ ]] = [[∃nas.[σ]iQ]]$, 
        \item[] $[[ [σ]∀pas.iN ]] = [[∀pas.[σ]iN]]$ (here we assume that $[[nas]]$ and $[[pas]]$ are lists of fresh variables, 
          that is the variable capture never happens);
      \end{itemize}
    \end{multicols}
\end{algorithm}

\begin{definition}[Substitution Signature]
  The signature $[[Γ' ⊢ σ : Γ]]$ means that
  \begin{enumerate}
    \item for any $[[α± ∊ Γ]], [[ Γ' ⊢ [σ]α± ]]$; and
    \item for any $[[α± ∉ Γ']], [[ [σ]α± = α± ]]$.
  \end{enumerate}
\end{definition}

A substitution can be restricted to a set of variables. 
The restricted substitution is define as expected. 
\begin{definition}[Subsitution Restriction]
  The specification $[[σ  | varset]]$ is defined as
  a function such that 
  \begin{enumerate}
    \item $[[σ|varset]]([[α± ]]) = [[σ]]([[α± ]])$, if $[[α± ]] \in [[varset]]$; and
    \item $[[σ|varset]]([[α± ]]) = [[α± ]]$, if $[[α± ]] \notin [[varset]]$.
  \end{enumerate}
\end{definition}

Two substitutions can be composed in two ways:
$[[σ2 ○ σ1]]$ corresponds to a consecutive application of $[[σ1]]$ and $[[σ2]]$,
while $[[σ2 <=< σ1]]$
depends on a signature of $[[σ1]]$ and modifies $[[σ1]]$ by applying
$[[σ2]]$ to its results on the domain.
\begin{definition}[Substitution Composition]
  $[[σ2 ○ σ1]]$ is defined as a function such that
  $[[σ2 ○ σ1]]([[α± ]]) = [[σ2]]([[σ1]]([[α± ]]))$.
\end{definition}

\begin{definition}[Monadic Substitution Composition]
  Suppose that $[[Γ' ⊢ σ1 : Γ]]$.
  Then we define $[[σ2 <=< σ1]]$ as $[[(σ2 ○ σ1)|Γ]]$.
\end{definition}
Notice that the result of $[[σ2 <=< σ1]]$ depends on the 
specification of $[[σ1]]$, which is not unique. 
However, we assume that the used specification clear from the 
context of the proof.

\begin{definition}[Equivalent Substitutions]
  The substitution equivalence judgement $[[Γ' ⊢ σ1 ≈ σ2 : Γ]]$ 
  indicates that on the domain $[[Γ]]$, 
  the result of $[[σ1]]$ and $[[σ2]]$ are equivalent in context $[[Γ']]$.
  Formally, for any $[[α± ∊ Γ]], [[ Γ' ⊢ [σ1]α± ≈ [σ2]α± ]]$.
\end{definition}

Sometimes it is convenient to construct substitution 
explicitly mapping each variable from a list (or a set)
to a type. Such substitutions are denoted as $[[iPs / pas]]$
and $[[iNs / nas]]$, where $[[iPs]]$ and $[[iNs]]$ are lists of 
the corresponding types.
\begin{definition}[Explicit Substitution]
  \hfill
  \begin{itemize}
    \item [$-$]
      Suppose that $[[nas]]$ is a list of negative type variables,
      and $[[iNs]]$ is a list of negative types of the same length.
      Then $[[iNs / nas]]$ denotes a substitution such that 
      \begin{enumerate}
        \item for $[[αi⁺ ∊ {nas}]]$, $[[ [iNs / nas] αi⁺]] = [[iNi]]$;
        \item for $[[β⁺ ∉ {nas}]]$, $[[ [iNs / nas] β⁺]] = [[β⁺]]$.
      \end{enumerate}
    \item [$+$]
      Positive explicit substitution $[[iPs / pas]]$
      is defined symmetrically.
  \end{itemize}
\end{definition}

\subsection{Declarative Subtyping}
Subtyping is one of the key mechanism of our system. 
It realizes the polymorphism: abstract $[[∀]]$ and 
$[[∃]]$ types can be used where concrete types are expected,
exactly because the subtyping relation between them.

\begin{definition} 
  \label{def:subDOne}
  \hfill

  \begin{multicols}{2}
    \ottdefnDOneNsub{}
    \columnbreak\\
    \ottdefnDOnePsup{}
  \end{multicols}

  \begin{multicols}{2}
    \ottdefnDOneNeq{}
    \columnbreak\\
    \ottdefnDOnePeq{}
  \end{multicols}
\end{definition}

The following observations about the declarative subtyping are worth noting:
\begin{itemize}
  \item \ruleref{\ottdruleDOneNVarLabel} and \ruleref{\ottdruleDOnePVarLabel}
    make the subtyping reflexive on variables (and further, on any type).
  \item \ruleref{\ottdruleDOneArrowLabel} is standard: the arrow is covariant on the
    resulting type and contravariant on the argument type.
  \item \ruleref{\ottdruleDOneShiftDLabel}  and \ruleref{\ottdruleDOneShiftULabel} are non-standard:
    the subtyping is \emph{invariant} for shifts. 
    This way, the subtyping of shifted types in one direction implies the subtyping
    in the opposite direction.
    Although this rule restricts the
    subtyping relation, it makes the system decidable.
  \item \ruleref{\ottdruleDOneForallLabel} and \ruleref{\ottdruleDOneExistsLabel} are the only
    non-algorithmic rules: the substitution for the quantified variable is
    not specified, those, these rules `drive' the subtyping relation.
\end{itemize}

In the next section, we present the sound and complete algorithm
checking whether one type is a subtype of another according to \cref{def:subDOne}.

\section{Algorithmic Type System}

\subsection{Grammar}

In the algorithmic system, we extend the grammar of types
by adding positive and negative \emph{algorithmic variables}
($[[α̂⁺]]$, $[[β̂⁺]]$, $[[γ̂⁺]]$, etc. and $[[α̂⁻]]$, $[[β̂⁻]]$, $[[γ̂⁻]]$, etc.).
They represent the unknown types, which will be inferred by the algorithm.
This way, we add two base cases to the grammar of 
positive and negative types, and use highlight to denote that the type
can potentially contain algorithmic variables.

\begin{definition}[Algorithmic Types]
  \label{def:algo-types}
  \hfill\\
  \begin{multicols}{2}
    \ottgrammartabular{
      \ottuN\ottinterrule
    }

    \ottgrammartabular{
      \ottuP\ottinterrule
    }
    \columnbreak
  \end{multicols}
\end{definition}

\subsection{Fresh Variable Selection}
\label{sec:fresh-selection}
Both the subtyping and the type inference algorithm
rely on the ability to select fresh, unused variables.
For a set of variables $[[varset]]$, it is indicated as 
$[[varset are fresh]]$ in the inference rules.
We assume that the selection subroutine always succeeds and is 
deterministic. In other words, whenever it is called in 
an algorithmic inference rule, it returns the same result, 
uniquely determined by the input of this rule.

\subsection{Variable Algorithmization}
\label{sec:variable-algorithmization}

In several places of our algorithm, in particular, during
algorithmic subtyping,
we turn a declarative type into the algorithmic one
via replacing certain type variables with fresh algorithmic variables.
We call this procedure \emph{variable algorithmization}, and define it as follows.

\begin{definition}[Variable Algorithmization]
  Suppose that $[[nas]]$ is a list of negative type variables
  and $[[nuas]]$ is a list of negative algorithmic variables of the same length. 
  Then $[[ nuas/nas ]]$ is a substitution-like procedure replacing each $[[αi⁻ ∊ {nas}]]$
  in a type for $[[αî⁻ ∊ {nuas}]]$.
\end{definition}

Conversely, we have the opposite procedure turning algorithmic type variables
into declarative type variables via \emph{dealgorithmization}.

\begin{definition}[Variable Dealgorithmization]
  Suppose that $[[nuas]]$ is a list of negative algorithmic variables
  and $[[nas]]$ is a list of negative type variables of the same length. 
  Then $[[ nas/nuas ]]$ is a substitution-like procedure replacing each
  $[[αî⁻ ∊ {nuas}]]$ in a type for $[[αi⁻ ∊ {nas}]]$.
\end{definition}

\subsection{Contexts and Well-formedness}

\begin{definition}[Algorithmic Type Context $[[Ξ]]$]
  \hfill \\
  Algorithmic type context $[[Ξ]]$ is represented by a set of 
  \emph{algorithmic} type variables ($[[α̂⁺]]$, $[[α̂⁻]]$, $[[β̂⁺]]$, \dots).
  The concatenation $[[Ξ1, Ξ2]]$ means the union of two contexts $[[Ξ1 ∪ Ξ2]]$.
\end{definition}

$[[Γ ; Ξ ⊢ uP]]$ and $[[Γ ; Ξ ⊢ uN]]$ are used to denote
that the algorithmic type is well-formed in the contexts
$[[Γ]]$ and $[[Ξ]]$, which means that each algorithmic variable
of the type is contained in $[[Ξ]]$, and each free declarative type variable
of the type is contained in $[[Γ]]$.

\begin{algorithm}[Algorithmic Type Well-formedness]
  \hfill

  \begin{multicols}{2}
  \ottdefnWFATNauWF{}
  \columnbreak

  \ottdefnWFATPauWF{}
  \end{multicols}

\end{algorithm}

Algorithmic Type Context are used in the unification algorithm.
In the subtyping algorithm, 
the context needs to remember additional information.
In the subtyping context, each algorithmic variable is associated with a
context it must be instantiated in 
(i.e. the context in which the type replacing the variable must be well-formed).
This association is represented by \emph{algorithmic subtyping context} $[[Θ]]$.
\begin{definition}[Algorithmic Subtyping Context $[[Θ]]$]
  \hfill \\
  Algorithmic Subtyping Context $[[Θ]]$ is represented by a set of 
  entries of form $[[ α̂⁺[Γ] ]]$ and $[[ α̂⁻[Γ] ]]$,
  where $[[α̂⁺]]$ and $[[α̂⁻]]$ are algorithmic variables,
  and $[[Γ]]$ is a context in which they must be instantiated.
  We assume that no two entries associating the same variable
  appear in $[[Θ]]$.

  $[[dom(Θ)]]$ denotes the set of variables appearing in $[[Θ]]$:
  $[[dom(Θ)]] = \{ [[α̂±]] \mid [[α̂±[Γ] ]] \in [[Θ]] \}$.
  If $[[ α̂±[Γ] ]] \in [[Θ]]$, we denote $[[Γ]]$ as $[[Θ(α̂±)]]$.
\end{definition}

\subsection{Subsitutions}

Substitution that operates on algorithmic type variables is denoted as
$[[uσ]]$. It is defined as a total function from algorithmic 
type variables to \emph{declarative} types, preserving the polarity.

The signature $[[Θ ⊢ uσ : Ξ]]$ means that
$[[Ξ ⊆ dom(Θ)]]$ and 
$[[uσ]]$ maps each algorithmic variable 
from $[[Ξ]]$ to a type well-formed in $[[Θ(α̂±)]]$;
and for each variable not appearing in $[[dom(Θ)]]$, 
it acts as identity.

\begin{definition}[Signature of Algorithmic Substitution]
  \label{def:algo-subst-sig}
  \hfill
  \begin{itemize}
    \item $[[Θ ⊢ uσ : Ξ]]$ means that
      \begin{enumerate}
        \item for any $[[α̂± ∊ Ξ]]$,
          there exists $[[Γ]]$ such that $[[ α̂±[Γ] ∊ Θ ]]$
          and $[[ Γ ⊢ [uσ]α̂± ]]$; 
        \item for any $[[ α̂± ∉ Ξ]]$, $[[ [uσ]α̂± ]] =  [[ α̂± ]]$.
      \end{enumerate}
    \item $[[Γ ⊢ uσ : Ξ]]$ means that
      \begin{enumerate}
        \item for any $[[α̂± ∊ Ξ]]$, $[[ Γ ⊢ [uσ]α̂± ]]$; 
        \item for any $[[ α̂± ∉ Ξ]]$, $[[ [uσ]α̂± ]] =  [[ α̂± ]]$.
      \end{enumerate}
  \end{itemize}
\end{definition}

In the anti-unification algorithm, we use another kind of substitution.
In contrast to algorithmic substitution $[[uσ]]$,
it allows mapping algorithmic variables to
\emph{algorithmic} types.
Additionally, anti-unification substitution is restricted to the
\emph{negative} segment of the language.
Anti-unification substitution is denoted as $[[aus]]$ and $[[ausr]]$.a

The pair of contexts $[[Γ]]$ and $[[Ξ]]$,
in which the results of an anti-unification substitution 
are formed, is fixed for this substitution.
This way, $[[Γ; Ξ2 ⊢ aus : Ξ1]]$ means that $[[aus]]$ maps each negative algorithmic
variable appearing in $[[Ξ1]]$ to a term well-formed in $[[Γ]]$ and $[[Ξ2]]$.

\begin{definition}[Signature of Anti-unification substitution]
  $[[Γ; Ξ2 ⊢ aus : Ξ1]]$ means that
  \begin{enumerate}
    \item for any $[[ α̂⁻ ∊ Ξ1]]$, $[[ Γ; Ξ2 ⊢ [aus]α̂⁻ ]]$ and
    \item for any $[[ α̂⁻ ∉ Ξ1]]$, $[[ [aus]α̂⁻ = α̂⁻ ]]$.
  \end{enumerate}
\end{definition}

\subsection{Equivalence and Normalization}
\label{sec:equivalence-normalization}

The subtyping-induced equivalence (\cref{def:subDOne}) is non-trivial:
there are types that are subtypes of each other but not equal. 
For example, $[[∀α⁺,β⁺.α⁺ → ↑β⁺]]$ is a subtype and a supertype of $[[∀α⁺,β⁺.β⁺ → ↑α⁺]]$
and of, for example, $[[∀α⁺,β⁺.β⁺ → ↑∃γ⁻.α⁺]]$, 
although these types are not alpha-equivalent.
For the subtyping algorithm, it is crucial to be able to check whether
two types are equivalent, without checking mutual subtyping. 
For this purpose we define the normalization procedure, 
which allows us to uniformly choose the representative type of the equivalence class.
This way, the equivalence checking is reduced to normalization and equality checking.

For clarification of the proofs and better understanding of the system, 
we introduce an intermediate relation---\emph{declarative equivalence}. 
As will be shown in \cref{lemma:equiv-soundness,lemma:equiv-completeness}, 
this relation is equivalent to the subtyping-induced equivalence, but does not 
depend on it. Although this relation is not defined algorithmically, 
it gives the intuition of what types our system considers equivalent.
Specifically, in addition to \emph{alpha-equivalence}, 
our system allows for \emph{reordering of adjacent quantifiers},
and \emph{introduction/elimination of unused quantifiers}.

The non-trivial rules of the declarative equivalence are
\ruleref{\ottdruleEOneForallLabel} and \ruleref{\ottdruleEOneExistsLabel}.
Intuitively, the variable bijection $[[μ]]$ reorders the quantifiers before
the recursive call on the body of the quantified type. 
It will be covered formally in \cref{sec:decl-equiv-lemmas}.

\begin{definition}[Declarative Type Equivalence]
  \hfill

  \begin{multicols}{2}
  \ottdefnEOneNeq{}
  \columnbreak\\
  \ottdefnEOnePeq{}
  \end{multicols}

\end{definition}

As the equivalence includes arbitrary reordering of quantified variables,
the normalization procedure is needed to choose the canonical order.
For this purpose, we introduce an auxiliary procedure---variable ordering. 
Intuitively, $[[ord varset in iN]]$ returns a list of variables from $[[varset]]$
in the order they appear in $[[iN]]$.

\begin{algorithm}[Variable Ordering]
  \label{alg:var-ordering}
  \hfill

  \begin{multicols}{2}
  \ottdefnONVar{}
  \columnbreak\\
  \ottdefnOPVar{}
  \end{multicols}

>>>>

\subsection{The Language}

@@
